# Setup {.unnumbered}

The `sits` package relies on the `sf` and `terra` **R** packages, which require the GDAL and PROJ libraries. Please follow the instructions below for installing `sf` and `terra` together with GDAL, provided by Edzer Pebesma.

## Support for GDAL and PROJ {.unnumbered}

### Windows and MacOS {.unnumbered}

Windows and MacOS users are strongly encouraged to install the `sf` and `terra` binary packages from CRAN. To install `sits` from source, please install package `Rtools` to have access to the compiling environment.

### Ubuntu {.unnumbered}

We recommend using the latest version of the GDAL, GEOS, and PROJ4 libraries and binaries. To do so, use the repository `ubuntugis-unstable`, which should be done as follows:

``` sh
sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable
sudo apt-get update
sudo apt-get install libudunits2-dev libgdal-dev libgeos-dev libproj-dev 
sudo apt-get install gdal-bin
sudo apt-get install proj-bin
```

Getting an error while adding this PPA repository could be due to the absence of the package `software-properties-common`. When GDAL is running in `docker` containers, please add the security flag `--security-opt seccomp=unconfined` on start. 

After installing GDAL, GEOS, and PROJ4, please install packages `sf` and `terra`, in this order.

```{r, eval = FALSE}
install.packages("sf")
install.packages("terra")
```


### Debian{-}

To install on Debian, use the [rocker geospatial](https://github.com/rocker-org/geospatial) dockerfiles. 

### Fedora {.unnumbered}

The following command installs all required dependencies:

``` sh
sudo dnf install gdal-devel proj-devel geos-devel sqlite-devel udunits2-devel
```

## Support for deep learning with torch {.unnumbered}

The deep learning models of `sits` use the `torch` package, which is an **R** version of `pyTorch`. Before installing `sits`, please also install packages `torch` and `luz`, and initialize torch. 

```{r, eval = FALSE}
install.packages("torch")
install.packages("luz")
torch::install_torch()
```


## Installing the `sits` package {.unnumbered}

After installing `sf`, `terra`, `torch`, and `luz`, please proceed to install `sits`,  which is available on CRAN and should be installed as a regular **R** package.

```{r, eval = FALSE}
install.packages("sits", dependencies = TRUE)
```

The source code repository is on [GitHub](https://github.com/e-sensing/sits). To install the development version of `sits`, which contains the latest updates but might be unstable, users should install `devtools`, if not already available, and then install `sits` as follows:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("e-sensing/sits@dev", dependencies = TRUE)
```

To run the examples in the book, please also install the `sitsdata` package.

```{r, eval = FALSE}
options(download.file.method = "wget")
devtools::install_github("e-sensing/sitsdata")
```


## Using GPUs with `sits` {.unnumbered}

The `torch` package automatically recognizes if a GPU is available on the machine and uses it for training and classification. There is a significant performance gain when GPUs are used instead of CPUs for deep learning models. There is no need for specific adjustments to `torch` scripts.  To use GPUs, `torch` requires version 11.6 of the CUDA library, which is available for Ubuntu 18.04 and 20.04. 
 
