# Setup {-}

The `sits` package relies on the `sf` and `terra` R packages, which in turn require the GDAL and PROJ libraries. Please follow the instructions below for installing `sf` and `terra` together with GDAL, provided by Edzer Pebesma. 

## Support for GDAL and PROJ{-}

#### Windows and MacOS{-}

Windows and MacOS users are strongly encouraged to install the `sf` and `terra` binary packages from CRAN. To install `sits` from source, please install package `Rtools` to have access to the compiling environment. 


#### Ubuntu{-}

For Unix-likes, GDAL (>= 2.0.1), GEOS (>= 3.4.0) and Proj.4 (>= 4.8.0) are required. Dependencies for recent versions of Ubuntu (18.04 and later) are available in the official repositories; you can install them with:

```sh
apt-get -y update && apt-get install -y  libudunits2-dev libgdal-dev libgeos-dev libproj-dev
```

To get more up-to-date versions of dependencies such as GDAL, we recommend adding the ubuntugis-unstable PPA to the package repositories and installing them as follows:

```sh
sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable
sudo apt-get update
sudo apt-get install libudunits2-dev libgdal-dev libgeos-dev libproj-dev 
```

Adding this PPA is required for installing `sf` on older versions of Ubuntu (e.g. Xenial).

#### Fedora{-}
The following command installs all required dependencies:
```sh
sudo dnf install gdal-devel proj-devel geos-devel sqlite-devel udunits2-devel
```

#### Arch{-}

Get gdal, proj and geos from the main repos, and udunits from the AUR:

```
pacman -S gdal proj geos
yay/pacaur/yaourt/whatever -S udunits
```

## Installing the `sits` package{-}

`sits` is currently available on GitHub <https://github.com/e-sensing/sits>. To install the package, install `devtools` if not already available, as then install sits as follows:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("e-sensing/sits", dependencies = TRUE)
```

To run the examples in the book, please also install the "sitsdata" package.
```{r, eval = FALSE}
devtools::install_github("e-sensing/sitsdata")
```