# Operations on data cubes{-}

```{r, echo = FALSE}
source("common.R")
library(sits)
```

## Pixel-based and neighborhood-based operations{-}

Pixel-based operations in remote sensing images refer to image processing techniques that operate on individual pixels or cells in an image without considering their spatial relationships with neighboring pixels. These operations are typically applied to each pixel in the image independently and can be used to extract information on spectral, radiometric, or spatial properties. Pixel-based operations produce spectral indexes which combine data from multiple bands.

Neighborhood-based operations are applied to groups of pixels in an image. The neighborhood is typically defined as a rectangular or circular region centered on a given pixel. These operations can be used for removing noise, detecting edges, and sharpening, among other uses.

The `sits_apply()` function computes new indices from a desired mathematical operation as a function of the bands available on the cube using any valid R expression. It applies the operation for all tiles and all temporal intervals. There are two types of operations in `sits_apply()`: 

1. Pixel-based operations that produce an index based on individual pixels of existing bands. The input bands and indexes should be part of the input data cube and have the same names used in the cube. The new index will be computed for every pixel of all images in the time series. Besides arithmetic operators, the function also accepts vectorized R functions that can be applied to matrices (e.g., `sqrt()`, `log()`, and `sin()`).

2. Neighborhood-based operations that produce a derived value based on a window centered around each individual pixel. The available functions  are `w_median()`, `w_sum()`, `w_mean()`, `w_min()`, `w_max()`, `w_sd()` (standard deviation), and `w_var()` (variance). Users set the window size (only odd values are allowed).

The following examples show how to use `sits_apply()`.

## Computing vegetation indexes{-}

Using vegetation indexes is an established practice in remote sensing. These indexes aim to improve the discrimination of vegetation structure by combining two wavebands, one where leaf pigments reflect incoming light with another where leaves absorb incoming radiation. Green leaves from natural vegetation such as forests have a strong emissivity rate in the near-infrared bands and low emissivity rates in the red bands of the electromagnetic spectrum. These spectral properties are used to calculate the Normalized Difference Vegetation Index (NDVI), a widely used index that is computed as the normalized difference between the values of infra-red and red bands. Including red-edge bands in Sentinel-2 images has broadened the scope of the bands used to calculate these indices [@Xie2019; @Sun2020a]. In what follows, we show examples of vegetation index calculation using a Sentinel-2 data cube. 

First, we define a data cube for a tile in the state of Rondonia, Brazil, including bands used to compute different vegetation indexes. We regularize the cube using a target resolution of 60 meters to reduce processing time. 

```{r, tidy="styler"}
# Create a directory to store files
if (!file.exists("./tempdir/chp5"))
    dir.create("./tempdir/chp5")
# Create an irregular data cube from MSPC
s2_cube <- sits_cube(
    source = "AWS",
    collection = "SENTINEL-S2-L2A-COGS",
    tiles = "20LKP",
    bands = c("B02", "B03", "B04", 
              "B05", "B06", "B07", 
              "B08", "B8A", "B11", 
              "B12","CLOUD"),
    start_date = as.Date("2018-07-01"),
    end_date = as.Date("2018-08-31"))
# Regularize the cube to 15 day intervals
reg_cube <- sits_regularize(
          cube       = s2_cube,
          output_dir = "./tempdir/chp5",
          res        = 60,
          period     = "P15D",
          multicores = 4)
```

There are many options for calculating vegetation indexes using Sentinel-2 bands. The most widely used method combines band B08 (785-899 nm) and band B04 (650-680 nm). Recent works in the literature propose using the red-edge bands B05 (698-713 nm), B06 (733-748 nm), and B07 (773-793 nm) for capturing subtle variations in chlorophyll absorption producing indexes, which are called Normalized Difference Vegetation Red-edge indexes (NDRE) [@Xie2019].  In a recent review, Chaves et al. [@Chaves2020] argue that red-edge bands are important for distinguishing leaf structure and chlorophyll content of different vegetation species. In the example below, we show how to include indexes in the regular data cube with the Sentinel-2 spectral bands. 

We first calculate the NDVI in the usual way, using bands B08 and B04.

```{r, tidy="styler", out.width="100%", fig.align="center", fig.cap="NDVI using bands B08 and B04 of Sentinel-2 (Source: Authors)."}

# Calculate NDVI index using bands B08 and B04
reg_cube <- sits_apply(reg_cube,
    NDVI = (B08 - B04)/(B08 + B04),
    output_dir = "./tempdir/chp5",
    multicores = 4,
    memsize = 12)
plot(reg_cube, band = "NDVI", palette = "RdYlGn")
```
We now compare the traditional NDVI with other vegetation index computed using red-edge bands. The example below such the NDRE1 index, obtained using bands B06 and B05. Sun et al. [@Sun2020a] argue that a vegetation index built using bands B06 and B07 provides a better approximation to leaf area index estimates than NDVI. Notice that the contrast between forests and deforested areas is more robust in the NDRE1 index than with NDVI. 

```{r, tidy="styler", out.width="100%", fig.align="center", fig.cap="NDRE1 using bands B06 and B05 of Sentinel-2 (Source: Authors)."}

# Calculate NDRE1 index using bands B06 and B05
reg_cube <- sits_apply(reg_cube,
    NDRE1 = (B06 - B05)/(B06 + B05),
    output_dir = "./tempdir/chp5",
    multicores = 4,
    memsize = 12)
# Plot NDRE1 index
plot(reg_cube, band = "NDRE1",  palette = "RdYlGn")
```


## Spectral indexes for identifying burned areas{-}

Band combination can also generate spectral indices for detecting degradation by fires, which are an important element in environmental degradation. Forest fires significantly impact emissions and impoverish natural ecosystems [@Nepstad1999]. Fires open the canopy, making the microclimate drier and increasing the amount of dry fuel [@Gao2020]. One well-established technique for detecting burned areas with remote sensing images is the normalized burn ratio (NBR), the difference between the near-infrared and the short wave infrared band,  calculated using bands B8A and B12.

```{r, tidy="styler", out.width="100%", fig.align="center", fig.cap="NBR ratio for a regular data cube built using Sentinel-2 tiles and 20LKP and 20LLP (Source: Authors)."}
# Calculate the NBR index
reg_cube <- sits_apply(reg_cube,
    NBR = (B12 - B8A)/(B12 + B8A),
    output_dir = "./tempdir/chp5",
    multicores = 4,
    memsize = 12)
# Plot the NBR for the first date"
plot(reg_cube, band = "NBR", palette = "Reds")
```

## Spectral mixture analysis{-}

Many pixels in images of medium-resolution satellites such as Landsat or Sentinel-2 contain a mixture of spectral responses of different land cover types inside a resolution element [@Roberts1993]. In many applications, it is desirable to obtain the proportion of a given class inside a mixed pixel. For this purpose, the literature proposes mixture models; these models represent pixel values as a combination of multiple pure land cover types [@Shimabukuro2019]. Assuming that the spectral response of pure land cover classes (called endmembers) is known, spectral mixture analysis derives new bands containing the proportion of each endmember inside a pixel.

The most used method for spectral mixture analysis is the linear model [@Shimabukuro2019]. The main idea behind the linear mixture model is that the observed pixel spectrum can be expressed as a linear combination of the spectra of the pure endmembers, weighted by their respective proportions (or abundances) within the pixel. Mathematically, the model can be represented as:
$$
R_i = \sum_{j=1}^N a_{i,j}*x_j + \epsilon_i, i \in {1,...M}, M > N,
$$
where $i=1,..M$ is the set of spectral bands and $j=1,..N$ is the set of land classes. For each pixel, $R_i$ is the reflectance in the i-th spectral band, $x_j$ is the reflectance value due to the j-th endmember, and $a_{i,j}$ is the proportion between the j-th endmember and the i-th spectral band. To solve this system of equations and obtain the proportion of each endmember, `sits` uses a non-negative least squares (NNLS) regression algorithm, which is available in the R package `RStoolbox` and was developed by Jakob Schwalb-Willmann, based on the sequential coordinate-wise algorithm (SCA) proposed on Franc et al. [@Franc2005]. 

To run the mixture model in `sits`, it is necessary to inform the values of pixels which represent spectral responses of a unique class. These are the so-called "pure" pixels. Because the quality of the resulting endmember images depends on the quality of the pure pixels, they should be chosen carefully and based on expert knowledge of the area.  Since `sits` supports multiple endmember spectral mixture analysis [@Roberts1998], users can specify more than one pure pixel per endmember to account for natural variability. 

In `sits`, spectral mixture analysis is done by `sits_mixture_model()`, which has two mandatory parameters: `cube` (a data cube) and `endmembers`, a named table (or equivalent) that defines the pure pixels. The `endmembers` table must have the following named columns: (a) `type`, which defines the class associated with an endmember; (b) names, the names of the bands. Each line of the table must contain the value of each endmember for all bands (see example). To improve readability, we suggest that the `endmembers` parameters be defined as a `tribble`. A `tribble` is a `tibble` with an easier to read row-by-row layout. In the example below, we define three endmembers for classes Forest, Soil, and Water.  Note that the values for each band are expressed as integers ranging from 0 to 10,000. 

```{r, tidy="styler", out.width="100%", fig.align="center", fig.cap="Percentage of forest per pixel estimated by mixture model (Source: Authors)."}
# Define the endmembers for three classes and six bands
em <- tibble::tribble(
    ~class,   ~B02, ~B03, ~B04, ~B8A, ~B11, ~B12,
    "forest",  200,  352,  189, 2800, 1340,  546,
    "soil",    400,  650,  700, 3600, 3500, 1800,
    "water",   700, 1100, 1400,  850,   40,   26)
# Generate the mixture model
reg_cube <- sits_mixture_model(
    data = reg_cube,
    endmembers = em,
    multicores = 4,
    memsize = 12,
    output_dir = "./tempdir/chp5")
# Plot the FOREST for the first date using the Greens palette
plot(reg_cube, band = "FOREST", palette = "Greens")
```

```{r, tidy="styler", out.width="100%", fig.align="center", fig.cap="Percentage of water per pixel estimated by mixture model (Source: Authors)."}
# Plot the water endmember for the first date using the Blues palette
plot(reg_cube, band = "WATER", palette = "Blues")
```

```{r, tidy="styler", out.width="100%", fig.align="center", fig.cap="Percentage of soil per pixel estimated by mixture model (Source: Authors)."}
# Plot the SOIL endmember for the first date using the orange red (OrRd) palette 
plot(reg_cube, band = "SOIL", palette = "OrRd")
```

Linear mixture models (LMM) improve the interpretation of remote sensing images by accounting for mixed pixels and providing a more accurate representation of the Earth's surface. Some of their key benefits include:

1. Improved classification accuracy: LMMs provide a more accurate representation of mixed pixels by considering the contributions of multiple land classes within a single pixel. This can lead to improved land cover classification accuracy compared to conventional per-pixel classification methods, which may struggle to accurately classify mixed pixels.

2. Sub-pixel information: LMMs allow for the estimation of the abundances of each land class within a pixel, providing valuable sub-pixel information. This can be especially useful in applications where the spatial resolution of the sensor is not fine enough to resolve individual land cover types, such as monitoring urban growth or studying vegetation dynamics.

3. Enhanced change detection: By considering the sub-pixel composition of land classes, LMMs can provide a more sensitive measure of changes in land cover over time. This can lead to more accurate and precise change detection, particularly in areas with complex land cover patterns or where subtle changes in land cover may occur.

4. Biophysical parameter estimation: LMMs can be used to estimate biophysical parameters, such as vegetation fraction, leaf area index (LAI), or soil moisture content, by relating the endmember abundances to these parameters. This can provide valuable information for monitoring and managing natural resources, agriculture, and ecosystems.

Applications of spectral mixture analysis in remote sensing include forest degradation [@Cochrane1998, @SouzaJr2005, @Shimabukuro2019, @Bullock2020, @Chen2021], wetland surface dynamics [@Halabisky2016], and urban area characterization [@Wu2003]. These models providing valuable information for a wide range of applications, from land mapping and change detection to resource management and environmental monitoring.

