# Example scripts

The following scripts show examples of how to use `sits` with increasing levels of complexity.

### Configuration to run the examples in this chapter{-}

```{r}
#| echo: false
reticulate::use_python("/Users/gilbertocamara/.pyenv/versions/pysits/bin/python")
```

:::{.panel-tabset}
## R
```{r}
#| echo: true
#| eval: true
#| output: false
# load package "tibble"
library(tibble)
# load packages "sits" and "sitsdata"
library(sits)
library(sitsdata)
# set tempdir if it does not exist
tempdir_r <- "~/sitsbook/tempdir/R/intro_examples"
if (!file.exists(tempdir_r))
  dir.create(tempdir_r, recursive = TRUE)
tempdir_cube <- paste0(tempdir_r, "/cube")
if (!file.exists(tempdir_cube))
  dir.create(tempdir_cube, recursive = TRUE)
tempdir_res <- paste0(tempdir_r, "/res")
if (!file.exists(tempdir_res))
  dir.create(tempdir_res, recursive = TRUE)
```
## Python
```{python}
#| echo: true
#| eval: false
#| output: false
from pysits import *
import pandas as pd
pd.set_option("display.max_columns", 100)
pd.set_option("display.max_rows", 4)
# set bookdir if it does not exist 
from pathlib import Path
home = str(Path.home())
tempdir_py = home + "/sitsbook/tempdir/Python/intro_examples"
Path(tempdir_py).mkdir(parents=True, exist_ok=True)
```
:::

## Cerrado classification

### Creating a data cube

```{r}
#| eval: true
# Define the roi for the samples_cerrado_lc8 dataset
roi_cerrado <- c(
    lon_min = -50.7809,
    lat_min = -13.3926,
    lon_max = -50.5402,
    lat_max = -13.2497)

# Define a data cube in the BDC repository
mpc_cube <- sits_cube(
    source = "MPC",
    collection  = "LANDSAT-C2-L2",
    bands = c("BLUE", "GREEN", "RED", "NIR08", "SWIR16", "SWIR22", "CLOUD"),
    roi = roi_cerrado,
    start_date = "2017-08-29",
    end_date = "2018-08-29"
)
```

### Exploring the features of the irregular cube

```{r}
#| echo: true
# Timeline
sits_timeline(mpc_cube)
```

```{r}
#| echo: true
# Bands
sits_bands(mpc_cube)
```

### Downloading cubo to local repository

```{r}
#| eval: false
#| echo: true
# Copy cube images to local directory
mpc_cube <- sits_cube_copy(
  cube = mpc_cube,
  roi = roi_cerrado,
  multicores = 5,
  output_dir = tempdir_cube
)
```

### Accessing data cube in local directory

```{r}
#| eval: true
#| echo: true
# Access local data cube
cube_local <- sits_cube(
    source = "MPC",
    collection = "LANDSAT-C2-L2",
    data_dir = tempdir_cube
)
```

### Regularizing Landsat data cube

```{r}
#| eval: false
#| echo: true
# Regularize
cube_reg <- sits_regularize(
    cube = cube_local,
    period = "P16D",
    res = 30,
    roi = roi_cerrado,
    multicores = 5,
    output_dir = tempdir_res,
    timeline = sits_timeline(samples_cerrado_lc8_examples),
    progress = TRUE
    )
```

```{r}
#| eval: true
#| echo: false
# Access local data cube
cube_reg <- sits_cube(
    source = "MPC",
    collection = "LANDSAT-C2-L2",
    data_dir = tempdir_res
)
```

### Exploring the features of the regular cube

```{r}
#| echo: true
# Timeline
sits_timeline(cube_reg)
```

```{r}
#| echo: true
# Bands
sits_bands(cube_reg)
```

### Computing indexes

```{r}
#| eval: false
#| echo: true
# Calculate NDVI index using bands NIR08 and RED
cube_reg <- sits_apply(
    data = cube_reg,
    NDVI = (NIR08 - RED) / (NIR08 + RED),
    output_dir = tempdir_res,
    multicores = 5,
    memsize    = 8,
    progress   = TRUE
  )

# Calculate EVI index using bands NIR08 and RED
cube_reg <- sits_apply(
    data       = cube_reg,
    EVI        = 2.5 * ((NIR08 - RED) / (NIR08 + 6 * RED - 7.5 * BLUE + 1)),
    output_dir = tempdir_res,
    multicores = 5,
    memsize    = 8,
    progress   = TRUE
    )
```

```{r}
#| echo: true
# Plot NDVI for the second date (2017-09-14)
plot(cube_reg,
  band = "NDVI",
  dates = "2017-09-14",
  palette = "RdYlGn"
)
```

### Getting time series from data cube

```{r}
#| eval: true
#| echo: true
# Extract time series
samples <- sits_get_data(
    cube = cube_reg,
    samples = samples_cerrado_lc8_examples,
    multicores = 5
)

# Show the tibble with the first three points
print(samples[1:3,])
```

```{r}
#| echo: true
# Check the sample distribution by class
summary(samples)
```

### Visualizing the spatial distribution of samples

```{r}
#| echo: true
# View samples
sits_view(samples)
```

### Colocar título

```{r}
#| echo: true
# Select a specific band from the cube
samples_ndvi <- sits_select(samples_cerrado_lc8_examples, bands = "NDVI")
plot(samples_ndvi)
```

```{r}
#| echo: true
# Calculate temporal patterns
patterns <- sits_patterns(samples_cerrado_lc8_examples)
plot(patterns)
```

### Fine-tuning the TempCNN model

```{r}
#|eval: false
#| echo: true
library(torch)
tuned_tempcnn <- sits_tuning(
    samples = samples_cerrado_lc8_examples,
    ml_method = sits_tempcnn(),
    params = sits_tuning_hparams(
        optimizer = torch::optim_adamw,
        opt_hparams = list(
            lr = uniform(10^-4, 10^-2),
            weight_decay = uniform(10^-6, 10^-3)
        )
    ),
    trials = 10,
    multicores = 8,
    progress = TRUE
)
```

```{r}
#|eval: true
#| echo: true
# Obtain accuracy, kappa, cnn_layers, cnn_kernels, and cnn_dropout_rates the best result:
tempcnn_parameters <- tuned_tempcnn[1, c("accuracy", "kappa", "cnn_layers", "cnn_kernels", "cnn_dropout_rates"), ]

# Organizing learning rates and weight decay in a list:
hparams_best <- tuned_tempcnn[1, ]$opt_hparams[[1]]

# Extracting learning rate and weight decay:'
lr_wd <- tibble::tibble(
    lr_best = hparams_best$lr,
    wd_best = hparams_best$weight_decay
)

# Printing the best parameters:
dplyr::bind_cols(tempcnn_parameters, lr_wd)
### Training a deep learning model
```

### Training a deep learning model

```{r}
#| eval: false
#| echo: true
library(torchopt)
set.seed(03022024)
# Train using tempCNN
tempcnn_model <- sits_train(
  samples_cerrado_lc8_examples,
  sits_tempcnn(
    optimizer            = torch::optim_adamw,
    cnn_layers           = c(256, 256, 256),
    cnn_kernels          = c(7, 7, 7),
    cnn_dropout_rates    = c(0.2, 0.2, 0.2),
    epochs               = 50,
    batch_size           = 64,
    validation_split     = 0.2,
    opt_hparams = list(lr = 0.006315829	,
                       weight_decay = 0.0003904277),
    verbose              = FALSE
  )
)

# Show training evolution
plot(tempcnn_model)
```

### Apply k-fold validation

```{r}
#| eval: false
#| echo: true
tempcnn_kfold <- sits_kfold_validate(
    samples = samples_cerrado_lc8_examples,
    folds = 5,
    ml_method = sits_tempcnn(                     
        optimizer = torch::optim_adamw,
        cnn_layers = c(256, 256, 256),
        cnn_kernels = c(7, 7, 7),
        cnn_dropout_rates = c(0.2, 0.2, 0.2),
        epochs = 50,
        batch_size = 64,
        validation_split = 0.2,
        opt_hparams = list(
        lr = 0.00578891,
        weight_decay = 0.00353351
        ),
        verbose = FALSE
    ),
    multicores = 7
)

tempcnn_kfold
```

### Apply F1-score validation

```{r}
#| eval: false
#| echo: true
tempcnn_f1score <- unname(tempcnn_kfold$byClass[, "F1"])

# F1-score result for TempCNN
tempcnn_f1score_acc <- tibble::tibble(
    "Classes" = sits_labels(samples_cerrado_lc8_examples),
    "TempCNN" = tempcnn_f1score
)

# Show the result
tempcnn_f1score_acc
```

### Data cube classification

```{r}
#| eval: false
#| echo: true
# Classify the raster image
cerrado_probs <- sits_classify(
  data = cube_local,
  ml_model = tempcnn_model,
  multicores = 5,
  memsize = 8,
  output_dir = tempdir_res
)
```

```{r}
#| eval: true
#| echo: true
# Plot the probability cube for class Forest
plot(cerrado_probs, labels = "Cerrado", palette = "BuGn")
```

### Spatial smoothing

```{r}
#| eval: false
#| echo: true
# Perform spatial smoothing
cerrado_bayes <- sits_smooth(
  cube = cerrado_probs,
  multicores = 5,
  memsize = 8,
  output_dir = tempdir_res
)
```

```{r}
#| eval: true
#| echo: true
plot(cerrado_bayes, labels = "Cerrado", palette = "BuGn")
```

### Labeling a probability data cube

```{r}
#| eval: false
#| echo: true
# Label the probability file
cerrado_map <- sits_label_classification(
  cube = cerrado_bayes,
  output_dir = tempdir_res
)
```

```{r}
#| eval: true
#| echo: true
plot(cerrado_map)
```

### Accuracy assessment of classified images

 \code{system.file("extdata/cerrado/validation.csv", package = "sitsdata")}

```{r}
#| eval: false
#| echo: true
# Get ground truth points
valid_csv <- system.file("extdata/csv/cerrado_lc8_validation.csv",
  package = "sitsdata"
)
# Calculate accuracy according to Olofsson's method
area_acc <- sits_accuracy(cerrado_map,
  validation = valid_csv,
  multicores = 4
)

# Print the area estimated accuracy
area_acc
```




