---
title: "Creating vector data cube from local files"
format: html
---

<a href="https://www.kaggle.com/code/esensing/creating-vector-data-cube-from-local-files" target="_blank">
<img src="https://kaggle.com/static/images/open-in-kaggle.svg"/>
</a>

### Configurations to run this chapter{-}

:::{.panel-tabset}
## R
```{r}
#| echo: true
#| eval: true
#| output: false
# Increase the timeout for download
options(timeout = 1200)
# load package "tibble"
library(tibble)
# load packages "sits" and "sitsdata"
library(sits)
library(sitsdata)
# set tempdir if it does not exist 
tempdir_r <- "~/sitsbook/tempdir/R/vec_create"
dir.create(tempdir_r, showWarnings = FALSE)
```
## Python
```{python}
#| echo: true
#| eval: false
#| output: false
# load "pysits" library
from pysits import *
from pathlib import Path
# set tempdir if it does not exist 
tempdir_py = Path.home() / "sitsbook/tempdir/Python/vec_create"
tempdir_py.mkdir(parents=True, exist_ok=True)
```
:::

## Introduction 

In many situations, users may want to create vector data cubes from local files. The most common case is when the want to recover a vector file has been previously created. An alternative is when users want to join a vector data file to a raster data cube. Recall that `sits` uses a specific notion of vector data cubes, which are polygons that cover an entire period of time. One example are farm boundaries for agricultural statistics. 

To be used for classification, vector cubes are associated to a raster cube. In this way, classification is done on a polygon-by-polygon basis. Given a polygon which matches part of a raster cube, `sits` uses pixels inside the polygon to obtain a set of time series that are used for classification, as shown in the example below.

In what follows, we illustrate the procedure of linking vector and raster file.


## Retrieving an external vector file 

In this example, we will use a vector file developed by the "Fields of the World" (FTW)
project [@Kerner2025]. FTW is a community-driven, global initiative that uses artificial intelligence and satellite imagery to map every agricultural field on Earth. Its aims are to detect and outline agricultural field boundaries at scale using machine learning and deep learning. It provide high-resolution boundary data spanning hundreds of countries and territories, mapping billions of individual field polygons. For more details in the project, please visit the [Fields of the World website](https://fieldsofthe.world/).

The selected FTW subset covers an agricultural area in the State of Bahia in Brazil and corresponds to an area inside the `S13W46` grid provide by FTW in their website, defined by the bounding box with limits `(lat_min = -11.82, lat_max = -11.57, lon_min = -45.75, lon_max = -45.52)`. The  FTW file contains closed polygons which are geometrically consistent. The data is available in the `sitsdata` package, as show below. 


:::{.panel-tabset}
## R
```{r}
#| label: list-fow-dir
#| echo: true
#| eval: true
#| cache: true
# directory where FTW file is located
fow_dir <- system.file("/extdata/fields_of_the_world", package = "sitsdata")
list.files(fow_dir)
```
## Python
```{python}
#| echo: true
#| eval: false
#| output: false
# FTW file is available in a local directory 
fow_dir = r_package_dir("extdata/Rondonia-20LMR/", package = "sitsdata")
# List FTW file
files = [p.name for p in Path(fow_dir).iterdir() if p.is_file()]
print(files)
```
```{r}
#| label: py-list-fow-dir
#| echo: false
#| eval: true
#| cache: true
fow_dir <- system.file("/extdata/fields_of_the_world", package = "sitsdata")
list.files(fow_dir)
```
:::

The FTW file has been converted from the original format `geoparquet`  to a `geopackage` format, which is used by `sits`. Its name has also been changed
to conform to the expected `sits` file format for raster and vector processed files, namely  `c("X1", "X2", "tile", "start_date", "end_date", "band", "version")`. In this case, the components of the file name are split using `_`, and the first two elements are ignored (although we have included `SENTINEL` and `FOW` for clarity). The band type is `segments`. Any geopackage polygon file that contains these required items can be imported and linked to a corresponding raster file, as shown below.


For further information, please see 
the section ["Reading classified images as a local data cube](https://e-sensing.github.io/sitsbook/dc_localcubes.html#reading-classified-images-as-local-data-cube) in this book. 

## Retrieving a local data cube

In this example, we will use an embedding cube covering the same region as the FTW segments. This cube has been produced using the scripts described in the ["Building regional embeddings"](https://e-sensing.github.io/sitsbook/emb_build.html). The original data consists of a two-year LANDSAT-8 cube for the period 2023-01-01 to 2024-12-31 covering the Brazilian Cerrado biome in monnthly intervals. To build the embedding, we used the JeLEPA algorithm [Balestriero2025] which is implemented by the `sits_ssl_lejepa()` function. 

:::{.panel-tabset}
## R
```{r}
#| label: fig-lejepa-cube
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|   RGB composite image for the LeJEPA embedding cube.
#| fig-align: center
#| out-width: 80%
# Retrieve a local embedding cube built with LeJEPA encoder
# data directory
images_dir <- system.file("extdata/Bahia-LeJEPA", package = "sitsdata")

# local cube
lejepa_cube <- sits_cube(
    source = "BDC",
    collection = "LANDSAT-OLI-16D",
    data_dir = images_dir,
    parse_info = c("satellite", "sensor", "tile", "band", "date")
)
# plot an RGB combination of three dimensions of the embedding cube
plot(lejepa_cube, green = "EMB10", red = "EMB12", blue = "EMB13")
```
## Python
```{python}
#| eval: false
# Retrieve a local cube based with MODIS data
# data directory
images_dir = r_package_dir("extdata/Bahia-LeJEPA", package = "sitsdata")

# local cube
lejepa_cube = sits_cube(
    source = "BDC",
    collection = "LANDSAT-OLI-16D",
    data_dir = images_dir,
    parse_info = c("satellite", "sensor", "tile", "band", "date")
)

# plot an RGB combination of three dimensions of the embedding cube
plot(lejepa_cube, green = "EMB10", red = "EMB12", blue = "EMB13")
```
```{r}
#| label: py-fig-lejepa-cube
#| echo: false
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|   RGB composite image for the LeJEPA embedding cube.
#| fig-align: center
#| out-width: 80%
# plot an RGB combination of three dimensions of the embedding cube
plot(lejepa_cube, green = "EMB10", red = "EMB12", blue = "EMB13")
```
:::

## Building a vector data cube for raster and vector files

We now produce a vector data cube by associating the FTW vector file
with the raster data cube produced above. To do this, we use the
`sits_cube()` function, with the following parameters:

| Parameter | Meaning |
|:----------|:--------|
| `source` | original source from which the data has been downloaded. |
| `collection` | image collection in data source. |
| `raster_cube` | raster cube to be merged with vector data. |
| `vector_dir` | local directory where vector files are stored. |
| `parse_info` | parsing information for metadata retrieval on local file. |
| `delim` | delimiter for parsing local files (default = "_"). |
| `multicores` | number of workers for parallel processing (default = 2). |
| `version` | version of the vector file (default = "v1"). |
| `progress` | show a progress bar? |

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-cube
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|   Vector data cube for a small area in the state of Bahia, Brazil.
#| fig-align: center
#| out-width: 80%
# create a vector cube by merging a raster cube with a vector file with segments
vec_lejepa_cube <- sits_cube(
    source = "BDC",
    collection = "LANDSAT-OLI-16D",
    raster_cube = lejepa_cube,
    vector_dir = fow_dir,
    parse_info = c("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot the segmented cube
plot(vec_lejepa_cube, 
     green = "EMB21", red = "EMB22", blue = "EMB23", 
     seg_color = "white", line_width = 0.6)
```
## Python
```{python}
#| echo: true
#| eval: false
#| label: py-noeval-fig-vec-cube
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|   Vector data cube for a small area in the state of Bahia, Brazil.
#| fig-align: center
#| out-width: 80%
# create a vector cube by merging a raster cube with a vector file with segments
vec_lejepa_cube = sits_cube(
    source = "BDC",
    collection = "LANDSAT-OLI-16D",
    raster_cube = lejepa_cube,
    vector_dir = fow_dir,
    parse_info = c("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot the segmented cube
plot(vec_lejepa_cube, 
     green = "EMB21", red = "EMB22", blue = "EMB23", 
     seg_color = "white", line_width = 0.6)
```
```{r}
#| label: py-fig-vec-cube
#| echo: false
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Vector data cube for a small area in the state of Bahia, Brazil.
#| fig-align: center
#| out-width: 80%
# plot the segmented cube
plot(vec_lejepa_cube, 
     green = "EMB21", red = "EMB22", blue = "EMB23", 
     seg_color = "white", line_width = 0.6)
```
:::

## Training a classification model 

After building a vector data cube, we can now classify it, following the guidelines 
of the previous chapter ["Object-based time series image analysis](https://e-sensing.github.io/sitsbook/vec_obia.html). We start by building 
a machine learning model using the MLP (multi-layer perceptron) algorithm. To do so,
we use a samples of a labelled training set developed by the `sits` team for LUCC classification in the Cerrado biome in Brazil. The data set is available in
Hugging Face. 

:::{.panel-tabset}
## R
```{r}
#| label: sample-dataset
#| echo: true
#| eval: true
#| cache: true
#| warning: false
# Download the labelled data set from Hugging Face 
samples_cerrado <- sits_from_hf(
  repo = "e-sensing/samples_cerrado",
  file = "labelled_samples_cerrado.parquet",
  type = "dataset"
)
# Use 10% of  samples
samples_cerrado_frac <- sits_sample(samples_cerrado, frac = 0.1)
# plot cube
summary(samples_cerrado_frac)
```
## Python
```{python}
#| echo: true
#| eval: false
# Download the labelled data set from Hugging Face 
samples_cerrado = sits_from_hf(
  repo = "e-sensing/samples_cerrado",
  file = "labelled_samples_cerrado.parquet",
  type = "dataset"
)
# Use 10% of  samples
samples_cerrado_frac = sits_sample(samples_cerrado, frac = 0.1)
# plot cube
summary(samples_cerrado_frac)
```
```{r}
#| label: py-sample-dataset
#| echo: false
#| eval: true
#| cache: true
#| warning: false
summary(samples_cerrado_frac)
```
:::

These samples were obtained by extracting time series for Landsat data cube. As 
such their data contains values for the original Landsat bands. To be used to 
train embeddings, they have to be converted to the same dimensions as those 
of embeddings cube. To do so, we follow the procedure used in the
["Building regional embeddings"](https://e-sensing.github.io/sitsbook/emb_build.html) chapter, which is to transform the samples using the same LeJEPA encoder that produced
the embeddings cube. For details on how to generate such encoder, please refer to
the above-mentioned chapter. For brevity's sake, we will use a version of the encoder which is archived in the Hugging Face repository. 

To transform the samples to the embedding space, we retrieve the LeJEPA encoder
from Hugging Face and use it to encode the samples.

:::{.panel-tabset}
## R
```{r}
#| label: sample-encoding
#| echo: true
#| eval: true
#| cache: true
#| cache.lazy: false
#| warning: false

# Download the encoder from Hugging Face 
lejepa_encoder <- sits_from_hf(
  repo = "e-sensing/encoders_cerrado", 
  file = "ssl_lejepa_tcnn_model_2017_2024.rds",
  type = "model"
)
# Encode the samples to transform them into the embedding space
samples_cerrado_enc <- sits_encode(
    data = samples_cerrado_frac, 
    encoder = lejepa_encoder)
```
## Python
```{python}
#| echo: true
#| eval: false
# Download the encoder from Hugging Face 
lejepa_encoder = sits_from_hf(
  repo = "e-sensing/encoders_cerrado", 
  file = "ssl_lejepa_tcnn_model_2017_2024.rds",
  type = "model"
)
# Encode the samples to transform them into the embedding space
samples_cerrado_enc = sits_encode(
    data = samples_cerrado_frac, 
    encoder = lejepa_encoder)
```
:::

After converting the samples to the embedding space, we build a classification model using the multi-layer perceptron algorithm (`sits_mlp()`). 

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-mlp
#| echo: true
#| eval: true
#| cache: true
#| cache.lazy: true
#| warning: false
#| fig-width: 7
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Accuracy and loss functions of the MLP model.
#| fig-align: center
#| out-width: 80%
# train mlp model
mlp_model <- sits_train(
    samples = samples_cerrado_enc,
    ml_method = sits_mlp(
        layers = c(256, 256, 256)
    )
)
# plot the evolution 
plot(mlp_model)
```
## Python
```{python}
#| eval: false
#| echo: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Evolution of accuracy and loss functions of the multi-layer perceptron model.
#| fig-align: center
#| out-width: 80%
# train mlp model
mlp_model = sits_train(
    samples = samples_cerrado_enc,
    ml_method = sits_mlp(
        layers = c(256, 256, 256)
    )
)
# plot the evolution 
plot(mlp_model)
```
```{r}
#| label: py-fig-vec-mlp
#| echo: false
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Evolution of accuracy and loss functions of the multi-layer perceptron model.
#| fig-align: center
#| out-width: 80%
# plot the evolution 
plot(mlp_model)
```
:::

## Obtaining a classified map using a vector data cube

To classify the cube `vector_lejepa_cube`, we follow the procedure described
in Chapter ["Object-based time series image analysis"](https://e-sensing.github.io/sitsbook/vec_obia.html). The first step
is a pixel-based classification using the trained model to the probabilities
associated to each pixel for each class. The `memsize`, `multicores` 
and `gpu_memory` parameters should be set by users according to the capabilities
of their machines.

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-local-probs
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Probabilites for class "Annual Crop"
#| fig-align: center
#| out-width: 80%
# classify the embeddings cube to obtain class probabilities
vec_lejepa_cube_probs <- sits_classify(
    data = vec_lejepa_cube,
    ml_model = mlp_model, 
    multicores = 4,
    memsize = 8, 
    gpu_memory = 8, 
    output_dir = tempdir_r
)
# plot the probabilities of label "Annual_Crop"
plot(vec_lejepa_cube_probs, labels = "Annual_Crop")
```
## Python
```{python}
#| eval: false
# classify the embeddings cube to obtain class probabilities
vec_lejepa_cube_probs = sits_classify(
    data = vector_lejepa_cube,
    ml_model = mlp_model, 
    roi = segments,
    multicores = 4,
    memsize = 8, 
    gpu_memory = 8, 
    output_dir = tempdir_py
)
# plot the probabilities of label "Annual_Crop"
plot(vec_lejepa_cube_probs, labels = "Annual_Crop")
```
```{r}
#| echo: false
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Probabilites for class "Annual Crop". 
#| fig-align: center
#| out-width: 80%
# plot the probabilities of label "Annual_Crop"
plot(vec_lejepa_cube_probs, labels = "Annual_Crop")
```
:::

After obtaning the probabilities, the next step is smoothing them. Given that
the input is a vector data cube, each segment defines a neighborhood boundary
for `sits_smooth()`. See the reference in the [previous chapter](https://e-sensing.github.io/sitsbook/vec_obia.html#example-of-snic-based-segmentation-and-classification). 

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-local-bayes
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Smoothed probabilites for class "Annual Crop". 
#| fig-align: center
#| out-width: 80%
# smoothen the probability cube
vec_lejepa_cube_bayes <- sits_smooth(
    cube = vec_lejepa_cube_probs,
    multicores = 4,
    memsize = 8, 
    gpu_memory = 8, 
    output_dir = tempdir_r
)
# plot the probabilities of label "Annual_Crop"
plot(vec_lejepa_cube_bayes, labels = "Annual_Crop")
```
## Python
```{python}
#| eval: false
# smoothen the probability cube
vec_lejepa_cube_bayes = sits_smooth(
    cube = vec_lejepa_cube_probs,
    multicores = 4,
    memsize = 8, 
    gpu_memory = 8, 
    output_dir = tempdir_py
)
# plot the smoothed probabilities of label "Annual_Crop"
plot(vec_lejepa_cube_bayes, labels = "Annual_Crop")
```
```{r}
#| echo: false
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Smoothed probabilites for class "Annual Crop".  
#| fig-align: center
#| out-width: 80%
# plot the smoothed probabilities of label "Annual_Crop"
plot(vec_lejepa_cube_bayes, labels = "Annual_Crop")
```
:::

The final step is producing a labelled classification map, based on 
the smoothed probability cube. For vector data cubes, 
`sits_label_classification()` performs segment-based labeling: 
pixel-level probabilities are aggregated inside each segment 
using the method specified by  the `label_method` parameter.
Options are "mean" (default), "median", or "majority".
All pixels within a segment receive the same class label in the output raster. 
The GPKG file with segments is updated to include a class column. 

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-local-class
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Labelled map for agricultural area in the state of Bahia, Brasil. 
#| fig-align: center
#| out-width: 90%
# label the vector cube with smoothed probabilities
vec_lejepa_cube_class <- sits_label_classification(
    cube = vec_lejepa_cube_bayes,
    label_method = "mean",
    multicores = 4,
    memsize = 8, 
    output_dir = tempdir_r
)
# plot the labelled map
plot(vec_lejepa_cube_class)
```
## Python
```{python}
#| eval: false
# label the vector cube with smoothed probabilities
vec_lejepa_cube_class = sits_label_classification(
    cube = vec_lejepa_cube_bayes,
    label_method = "mean",
    multicores = 4,
    memsize = 8, 
    output_dir = tempdir_py
)
# plot the labelled map
plot(vector_lejepa_cube_class)
```
```{r}
#| echo: false
#| cache: true
#| warning: false
#| fig-width: 8
#| fig-height: 5
#| fig-dpi: 300
#| fig-cap: |
#|  Labelled map for agricultural area in the state of Bahia, Brasil.
#| fig-align: center
#| out-width: 90%
# plot the labelled map
plot(vec_lejepa_cube_class)
```
:::

## Summary

In this chapter, we show how mix a vector file with a raster data cube to build
a vector data cube. The vector file must be in the geopackage format and its polygons are contained inside the extent of the raster data cube. Also, the vector file 
name needs to provide adequate metadata. This allows much flexibility to users to combine external polygon files to sits. 
