---
title: "Creating vector data cube from local files"
format: html
---

<a href="https://www.kaggle.com/code/esensing/creating-vector-data-cube-from-local-files" target="_blank">
<img src="https://kaggle.com/static/images/open-in-kaggle.svg"/>
</a>

### Configurations to run this chapter{-}

:::{.panel-tabset}
## R
```{r}
#| echo: true
#| eval: true
#| output: false
# load package "tibble"
library(tibble)
# load packages "sits" and "sitsdata"
library(sits)
library(sitsdata)
# set tempdir if it does not exist 
tempdir_r <- "~/sitsbook/tempdir/R/vc_create"
dir.create(tempdir_r, showWarnings = FALSE)
```
## Python
```{python}
#| echo: true
#| eval: false
#| output: false
# load "pysits" library
from pysits import *
from pathlib import Path
# set tempdir if it does not exist 
tempdir_py = Path.home() / "sitsbook/tempdir/Python/vc_create"
tempdir_py.mkdir(parents=True, exist_ok=True)
```
:::

## Introduction 

In many situations, users may want to create vector data cubes from local files. The most common case is when the want to recover a vector file has been previously created. An alternative is when users want to join a vector data file to a raster data cube. Recall that `sits` uses a specific notion of vector data cubes, which are polygons that cover an entire period of time. One example are farm boundaries for agricultural statistics. 

To be used for classification, vector cubes are associated to a raster cube. In this way, classification is done on a polygon-by-polygon basis. Given a polygon which matches part of a raster cube, `sits` uses pixels inside the polygon to obtain a set of time series that are used for classification, as shown in the example below.

In what follows, we illustrate the procedure of linking vector and raster file.

## Retrieving a local data cube

We take the MODIS data set for a small area in Mato Grosso, Brazil available in the `sits` package. For details on how to retrieve raster data cubes from local files, please see chapter "Data cubes from local files". 

:::{.panel-tabset}
## R
```{r}
#| label: vec-modis-cube
#| eval: true
#| cache: true
#| warning: false
# Retrieve a local cube based with MODIS data
# data directory
data_dir <- system.file("extdata/raster/mod13q1", package = "sits")

# local cube
modis_cube <- sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    data_dir = data_dir,
    parse_info = c("satellite", "sensor", "tile", "band", "date")
)
```
## Python
```{python}
#| eval: false
# Retrieve a local cube based with MODIS data
# data directory
data_dir = r_package_dir("extdata/raster/mod13q1", package = "sits")

# local cube
modis_cube = sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    data_dir = data_dir,
    parse_info = ("satellite", "sensor", "tile", "band", "date")
)
```
:::

Then, we produce a vector data cube associated to the raster data cube using `sits_segment`, as described in the previous chapter.

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-modis-segments
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|   Example of MODIS image segmentation for a small area in Sinop, Brazil.
#| fig-align: center
#| out-width: 80%
# segment the vector cube
segs_cube <- sits_segment(
    cube = modis_cube,
    output_dir = tempdir_r 
)

# plot the segmented cube
plot(segs_cube)
```
## Python
```{python}
#| echo: true
#| eval: false
# segment the vector cube
segs_cube = sits_segment(
    cube = modis_cube,
    output_dir = tempdir_py
)

# plot the segmented cube
plot(segs_cube)
```
```{r}
#| label: py-fig-vec-modis-segments
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Example of MODIS image segmentation for a small area in Sinop, Brazil..
#| fig-align: center
#| out-width: 80%
# plot the segmented cube
plot(segs_cube)
```
:::

The resulting vector file is stored in the `tempdir` directory, in the `geopackage` format. The file should contain closed polygons which are geometrically consistent. Listing the resulting file, one can better understand the convention.

:::{.panel-tabset}
## R
```{r}
#| echo: true
#| eval: true
#| cache: true
list.files(tempdir_r, pattern = "*segments*")
```
## Python
```{python}
#| echo: true
#| eval: false
list(
    tempdir_py.glob(pattern = "*segments*")
)
```
```{r}
#| echo: false
#| eval: true
#| cache: true
list.files(tempdir_r, pattern = "*segments*")
```
:::

The segments file is a geopackage whose title contains optional information on satellite and sensor  and mandatory information on `tile`, `start_date`, `end_date`, `band` (in this case `segments`) and `version`. Any geopackage polygon file that contains these required items can be imported and linked to a corresponding raster file, as shown below. 

## Linking raster and vector data 

Now consider the inverse situation, where a set of a segments is already stored locally and users want to recover them and associate to a raster data cube. In the same way as we do for raster data cubes in [Data cubes from local files](https://e-sensing.github.io/sitsbook/dc_localcubes.html), we use the `sits_cube()` function with the following parameters:

- `source`: data source of the raster data (in this case, the Brazil Data Cube).
- `collection`: ARD collection associated to the images (in the example, `MOD13Q1-6.1").
- `raster_cube`: raster data cube to associate with the polygons.
- `vector_dir`: directory where the vectors are stored. 
- `parse_info`: how to extract metadata from the name of the `geopackage` file. The minimum required elements are `tile`, `start_date`, `end_date`, `band` and `version`. The start and end dates must match those of the raster cube.
- `delim`: delimiter character for breaking down the file into parts ("_" by default.)

To import an external `geopackage` file, it must be named according to a convention that is understable by `sits`. The file needs to contain the required information (`tile`, `start_date`, `end_date`, `band` and `version`) separated by underlines. All other parts of the `geopackage` file name can use dummy values (e.g, `X1`). In the example below, we recover the file we produced before. We could replace it by any other polygon file in geopackage format that matches the `parse_info` information listed above. As `satellite` and `sensor` are optional, we use place holders from them.

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-local-segments
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Vector data cube which combines segments and raster files.
#| fig-align: center
#| out-width: 80%
# recover the local segmented cube
local_segs_cube <- sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = modis_cube,
    vector_dir = tempdir_r,
    parse_info =  c("X1", "X2", "tile", "start_date", "end_date", "band", "version"),
    delim = "_"
)

# plot the recovered model and compare with previous plot
plot(local_segs_cube, line_width = 1.0)
```
## Python
```{python}
#| eval: false
# recover the local segmented cube
local_segs_cube = sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = modis_cube,
    vector_dir = tempdir_py,
    vector_band = "segments",
    parse_info = ("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot the recover model and compare with previous plot
plot(local_segs_cube)
```
```{r}
#| label: py-fig-vec-local-segments
#| echo: false
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Vector data cube which combines segments and raster files.
#| fig-align: center
#| out-width: 80%
# plot the recovered model and compare with previous plot
plot(local_segs_cube, line_width = 1.0)
```
:::

The resulting plot should be the same as the original one. We now train a model with the Random Forest algorithm, and use the result to classify the segments. Then we show how to recover the classified data from the local files which was produced by the classifier. 

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-probs-cube
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Probabilities of class Forest in the classified vector data cube.
#| fig-align: center
#| out-width: 80%
# create a Random Forest model
rfor_model <- sits_train(samples_modis_ndvi, sits_rfor())
# classify the segments
probs_vector_cube <- sits_classify(
    data = segs_cube,
    ml_model = rfor_model,
    output_dir = tempdir_r,
)
# plot cube
plot(probs_vector_cube, labels = "Forest")
```
## Python
```{python}
#| echo: true
#| eval: false
# classify the segments
# create a Random Forest model
rfor_model = sits_train(samples_modis_ndvi, sits_rfor())
probs_vector_cube = sits_classify(
    data = segs_cube,
    ml_model = rfor_model,
    output_dir = tempdir_py,
    n_sam_pol = 10
)

# plot cube
plot(probs_vector_cube, labels = "Forest")
```
```{r}
#| label: py-fig-vec-probs-cube
#| echo: false
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Probabilities of class Forest in the classified vector data cube.
#| fig-align: center
#| out-width: 80%
# plot cube
plot(probs_vector_cube, labels = "Forest")
```
:::

In the same way as before, we can recover the probability cube from local files using `sits_cube()`.

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-probs-recover
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Probability cube recovered from local raster cube and segments files.
#| fig-align: center
#| out-width: 80%
# recover vector cube
local_probs_vector_cube <- sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = probs_vector_cube,
    vector_dir = tempdir_r,
    parse_info =  c("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot
plot(local_probs_vector_cube, labels = "Forest")
```
## Python
```{python}
#| eval: false
# recover vector cube
local_probs_vector_cube = sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = probs_vector_cube,
    vector_dir = tempdir_py,
    parse_info =  ("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot
plot(local_probs_vector_cube, labels = "Forest")
```
```{r}
#| label: py-fig-vec-probs-recover
#| echo: false
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Probability cube recovered from local raster cube and segments files.
#| fig-align: center
#| out-width: 80%
# recover vector cube
local_probs_vector_cube <- sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = probs_vector_cube,
    vector_dir = tempdir_r,
    parse_info =  c("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot
plot(local_probs_vector_cube, labels = "Forest")
```

:::

Finally, we produce a classified map.

:::{.panel-tabset}
## R
```{r}
#| label: fig-vec-local-class
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Labeled vector data cube based on the probabilities.
#| fig-align: center
#| out-width: 80%
# label the segments
class_vector_cube <- sits_label_classification(
    cube = probs_vector_cube,
    output_dir = tempdir_r,
    label_method = "mean"
)
plot(class_vector_cube)
```
## Python
```{python}
#| eval: false
# label the segments
class_vector_cube = sits_label_classification(
    cube = probs_vector_cube,
    label_method = "mean",
    output_dir = tempdir_py,
)

# plot
plot(class_vector_cube)
```
```{r}
#| label: py-fig-vec-local-class
#| echo: true
#| eval: true
#| cache: true
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 300
#| fig-cap: |
#|  Labeled vector data cube based on the probabilities.
#| fig-align: center
#| out-width: 80%
# plot
plot(class_vector_cube)
```
:::

We then show an example of how to recover a labeled vector data cube from local files. As in the previous examples, the segments file is located in the `vector_dir`, and the parameter `parse_info` tells `sits` how to recover its metadata.

:::{.panel-tabset}
## R
```{r}
# recover vector cube
local_class_vector_cube <- sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = class_vector_cube,
    vector_dir = tempdir_r,
    parse_info =  c("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot
plot(local_class_vector_cube)
```
## Python
```{python}
# recover vector cube
local_class_vector_cube = sits_cube(
    source = "BDC",
    collection = "MOD13Q1-6.1",
    raster_cube = modis_cube,
    vector_dir = tempdir_py,
    vector_band = "class",
    parse_info =  ("X1", "X2", "tile", "start_date", "end_date", "band", "version")
)

# plot
plot(local_class_vector_cube)
```

:::

## Summary

In this chapter, we show how mix vector and raster data cubes. For the operation to succeed, the vector cube must be a geopackage file whose polygons and contained inside the raster data cube and files need to provide adequate metadata. The polygon file can come from outside sources, provided it spatio-temporal coordinates match those of the raster data cube to which it is being merged. This allows much flexibility to users to combine external polygon files to sits. 
